---
title: "IFR Flows"
author: "Enrico Spinielli"
date: "August 19, 2016"
output: 
  pdf_document: 
    highlight: haddock
    number_sections: yes
    toc: yes
documentclass: article
classoption: a4paper
bibliography: biblio.bib
---
# Abstract

This document describes the assumptions, shortcuts and steps taken to prepare the
data used the circular plots of the flow of flights entering/exiting and within
the airspace of Eurocontrol's Member States for the years 2001-2015 as displayed
on PRU's website `[`[1][extra]`]`, `[`[2][intra]`]`.

# Introduction

The studies for the [flow of flights entering/exiting][extra] and [within][intra]
the airspace of Eurocontrol's Member States is inspired by Sander, Abel and
Bauer's website [see @global-refugees] and its companion working paper
[see @glob-migration-viz].

The underlying dataset is coming from the PRU's Analytical database
and takes into account all [IFR flights][ifr] in the period 2001-2016
as recorded by Eurocontrol's [Network Manager][nm] (formely [CFMU][cfmu]).

In the following chapters we will describe how we extracted the flows, what we filtered,
the amount and impact of the filters, the compromises we have taken when mapping
flights to countries and countries to regions and more.

This is a work in progress so throughout this document we will mark what needs
to be tackled soon[^soon] with `TODO` and areas of further
investigation[^deeper] with `RSRC`.

[^soon]: The definition of *soon* is intentionally omitted to let us free to decide ;-)

[^deeper]: These are subjects that could potentially (typically because of our lack
  of knowledge) require a substantial amount of time/resources.


# The (nasty) details

## Yearly city pair flows 

The following query extracts city pair flows from the archived flight details in PRU's
analytical database[^fp].

[^fp]: We have access to flight details (and more) since $27^{th}$ Apr 1995
  when the flight plan reception and processing for France, Germany and the
  Benelux States was handled by the then new Central Flowm Management Unit.

```{mysql, eval=FALSE}
set worksheetname Flows;
SET SQLFORMAT csv;

SELECT
  TO_CHAR(F.LOBT, 'YYYY') AS YEAR
  , COUNT(FLT_UID) NB_OF_FLIGHT
  , A1.EC_ISO_CT_CODE DEP_ISO_COUNTRY_CODE
  , A2.EC_ISO_CT_CODE DES_ISO_COUNTRY_CODE
FROM
    SWH_FCT.FAC_FLIGHT F
  , SWH_FCT.DIM_AIRPORT A1
  , SWH_FCT.DIM_AIRPORT A2
  , SWH_FCT.DIM_AIRCRAFT_TYPE AC
WHERE
    A1.SK_AP_ID = F.SK_ADEP_ID
  AND
    A2.SK_AP_ID   = F.SK_ADES_ID
  AND
    F.LOBT       >= '01-JAN-2001'
  AND
    F.LOBT        < '01-JAN-2016'
  AND
    (A1.EC_ISO_CT_CODE != '##' AND A2.EC_ISO_CT_CODE != '##')
  AND
    F.AIRCRAFT_TYPE_ICAO_ID = AC.ICAO_TYPE_CODE
  AND
    (F.ICAO_FLT_TYPE != 'M')
  AND
    SUBSTR(AC.ICAO_DESC,1,1) != 'H'
GROUP BY
    A1.EC_ISO_CT_CODE
  , A1.ISO_CT_CODE
  , A2.EC_ISO_CT_CODE
  , TO_CHAR(F.LOBT,'YYYY');
```

The flollowing filters have been applied in the query above:

* flights without an ISO2 country code for either the departing or arriving airport have been excluded:
  
    ```{mysql, eval=FALSE}
    ...
    AND (A1.EC_ISO_CT_CODE != '##' AND A2.EC_ISO_CT_CODE != '##')
    ...
    ```

* military and helicopter flights have been excluded:

    ```{mysql, eval=FALSE}
    ...
    AND (F.ICAO_FLT_TYPE != 'M')
    AND SUBSTR(AC.ICAO_DESC,1,1) != 'H'
    ...
    ```


The following table summarises the amount of filtered out flights compared to the total per year.

| Year | No ISO code | Military | Helicopter |
|------|-------------|----------|------------|


* **TODO** fill the table above with the result of the queries we run in the past
  in order to justify the next paragraph.

Given the filtered out amounts are tiny compared with the relevant totals, it is
justified to discard them because they are not going to distort the shape of the flows
once we aggregate the flights per country and region.



## Preprocessing

The raw data is preprocessed via the script `preprocessRawData.R`

```{r cache=FALSE, echo=FALSE, tidy.opts=list(blank=FALSE, width.cutoff=50)}
knitr::read_chunk('R/preprocessRawData.R')
```

We define the range of years we will process:
```{r configuration, eval=FALSE}
```

and filter the dataset accordingly:
```{r filter-years, eval=FALSE}
```


Then columns names are redefined:
```{r rename-columns, eval=FALSE}
```

Let's make sure the dataset is sound:

* `year` is within the 2001 - 2015 range
* origin and destinations are not undefined, (but "NA" is a valid value for Namibia)

```{r check-invariants, eval=FALSE}
```

Also due to our mapping of old ISO2 codes from our DB to more *modern* ones from the relevant year.
This is the case for Serbia (`CS` to `RS`) and Martinique/Guadaloupe (`XF` to `MQ`)

```{r fix-iso-codes, eval=FALSE}
```

* `TODO` check ISO codes for Montenegro (ME since 2006), Serbia (RS since 2006),
  Congo (CD, since 1997), YU (Yugoslavia, till 2003-07), ZR (Zaire, till 1995-07)

* `TODO`: check what happens for
  [Dutch Antilles](https://en.wikipedia.org/wiki/Netherlands_Antilles),
  France's territories outside Europe (St Pierre's aiport,
  LFVP, in the oversea community of Saint-Pierre and Miquelon; or
  [the ones in the French Guiana](https://en.wikipedia.org/wiki/List_of_airports_in_French_Guiana),
  ...)

* `RSRC` check how ISO 3166-2 codes evolved in time and how that
  is handled in the `EC_ISO_CT_CODE` column of the `SWH_FCT.DIM_AIRPORT` table.
  The [relevant page on Wikipedia][iso2c] provides some info on the time when
  codes changed.

* `RSRC` if we are going to map the flows, it could be interesting to try to find
  shapefiles for how countries did split, reshaped, ...


Finally we save to output filename for further use.

```{r file-names, eval=FALSE}
```
```{r save-preprocessed, eval=FALSE}
```

```{r eval=FALSE, echo=FALSE}
source("R/preprocessRawData.R")
```



## Data transformation for circular plot

The circular plot clusters countries into regions.
Luckily the [`countrycode`][ccode] R package converts country codes into
one of eleven coding schemes including regional grouping.
A recently added feature allows to provide a dictionary for custom mappings:
we can hence use one for the our aviation data[^ccavia].

[^ccavia]: the README from the package repository describes this feature
  using a dataset we provided as a [contribution to the package][ccodepr]

```{r cache=FALSE, echo=FALSE, tidy.opts=list(blank=FALSE, width.cutoff=50)}
knitr::read_chunk('R/mapToRegions.R')
```


### Preliminary fixes

We start by reading the previously processed data,

```{r read-preprocessed-data, eval=FALSE}
```


We then (arbitrarily) assign a UN region to countries that do not have one,
i.e. Taiwan (ISO 3166-2: `TW`) and 
the British Indian Ocean Territory (ISO 3166-2: `IO`).

```{r fix-missing-un-region, eval=FALSE}
```

### Assign region to country

Assign regions as per Eurocontrol's PRU classification

| PRU's region | Countries   |
|--------------|-------------|
|              |             |

* `TODO`: generate a table with the mapping, possibly using `kable`,
  see [here](http://stackoverflow.com/a/19859769/963575)


```{r classify-pru-regions, eval=FALSE}
```


### Split dataset: intra and extra Eurocontrol
We have 3 classes of flows:

1. Flows within Eurocontrol area, i.e. flights where both departure and destination airports are
   in Eurocontrol Member States's territory.
1. Overfling flows, i.e. flights originating and landing in non-Member States
1. Flows in/out of Eurocontrol area, i.e. flights where either departure or destination airport is
   in a Member State.

Using the following tables (IN = inside Eurocontrol, OUT = outside Eurocontrol)

Departure AD | Destination AD | Flow type       | Description
-------------|----------------|-----------------|------------------
  IN         |  IN            | internal        | Internal flows
  IN         |  OUT           | outbound        | Outflying flows
  OUT        |  IN            | inbound         | Inflying flows
  OUT        |  OUT           | overflying      | Overflying flows

we can classify the flows

```{r classify-flows, eval=FALSE}
```

We now split the dataset in two: `intra-flows` contains all `internal` flows;
`extra-flows` all the rest (i.e. we will not separate the overflying flow from
In/Outflying ones.)

```{r intra-flows, eval=FALSE}
```

Furthermore we map `intra-flows` regions according to ESRA08 classification
from STATFOR [see pag 57 of @statfor-medium-2016]:
```{r classify-intra-statfor, eval=FALSE}
```

Save the internal and external flows on disk:

```{r save-intra-flows, eval=FALSE}
```


### Refine Extra Flows

In order to reduce the number of regions in the final circular plot,
we reassign the regions to some extra-European flows.
So `Russia` and `China` go to `Asia` rather than having their own region.
We also merge the whole Northern, Central and Souther America in one `America` region.

```{r extra-flow-regions-remapping, eval=FALSE}
```

Save the external flows on disk:
```{r save-extra-flows, eval=FALSE}
```

```{r eval=FALSE, echo=FALSE}
source("R/mapToRegions.R")
```



## Analysis

* `RSRC`: analyse the flows time series.
Summaries:

```{r, echo=FALSE, eval=FALSE}
extra_stats <- extra_flows %>%
  mutate(flow_name = ifelse(originregion_name < destinationregion_name, 
                            paste0(originregion_name, "-", destinationregion_name),
                            paste0(destinationregion_name, "-", originregion_name))) %>%
  select(year, flights, flow_name) %>%
  group_by(year, flow_name) %>%
  summarise(movements=sum(flights)) %>%
  arrange(desc(movements)) %>%
  spread(year, movements, fill = 0)
```

List of countries contributing the major movements:

```{r, echo=FALSE, eval=FALSE}
# limit to 2014 (for now)
extra_movements <- extra_flows %>% 
  # exclude overflying flows
  filter(originregion_name == 'Eurocontrol' |
           destinationregion_name == 'Eurocontrol') %>%
  select(origin_name, destination_name,
         originregion_name,
         destinationregion_name,
         countryflow_2014) %>%
  mutate(flow = ifelse(originregion_name != 'Eurocontrol',
                       origin_name,
                       destination_name))

# print 30 biggest contributors
extra_movements %>% 
  group_by(flow) %>% 
  summarise(total = sum(countryflow_2014)) %>%
  arrange(desc(total)) %>%
  print(n=30)
```








## Data Preparation for Visualisation 

The [circular plot library[^gmlib]][gmrepo] we use for visualizing the flows of flights
expects as input preprocessed data ready to plot.

[^gmlib]: This library is an extension of the [chord diagram][chorddia]
  as made possible by Mike Bostock's [D3.js][d3js].

In this section we explain which are the transformations and support
files needed. These support files will be *compiled* by the code
in another repository, [`circflows`][circflows].

We will pick the extra flow dataset but the same set of files is
produced for the intra flow one.


```{r cache=FALSE, echo=FALSE, tidy.opts=list(blank=FALSE, width.cutoff=50)}
knitr::read_chunk('R/prepareForVizExtra.R')
```


### Countries to show

In order to avoid too small flows in the final visualization we filter out
all country flows with less that 10 flights per day, i.e. $10 * 365$.

```{r threshold-extra, eval=FALSE}
```

Then we will then generate a file that specifies which countries to show, something like:

```{R, eval=FALSE}
iso,show (1 = yes; 2 = no)
US,1
IN,1
CN,2
...
```


Now let's filter out the little flows...

```{r extra-viz-thresholded, eval=FALSE}
```

### Years in the dataset

We need a file listing the years in the dataset

```{r create-years-file, eval=FALSE}
```

* `TODO`: generate years file for intra-flow too.


### Countries and regions in the datasets

We then need to create the files for the countries and the regions present
in the (now filtered) dataset

```{r create-countries-file, eval=FALSE}
```

```{r create-regions-file, eval=FALSE}

```


```{r eval=FALSE, echo=FALSE}
source("R/prepareForVizExtra.R")
```

```{r eval=FALSE, echo=FALSE}
source("R/prepareForVizIntra.R")
```


# References


[cfmu]: <http://www.eurocontrol.int/20-years-central-flow-management> "Central Flow Management - The twenty-year journey"
[nm]: <https://www.eurocontrol.int/network-manager> "Network Manager"
[extra]: <http://ansperformance.eu/studies/flows-extra/> "Flows of flights in/out of Europe"
[intra]: <http://ansperformance.eu/studies/flows-intra/> "Flows of flights within of Europe"
[iso2c]: <https://en.wikipedia.org/wiki/ISO_3166-1_alpha-2> "ISO 3166-1 alpha-2"
[ccode]: <https://github.com/vincentarelbundock/countrycode> "R package countrycode"
[ccodepr]: <https://github.com/vincentarelbundock/countrycode/pull/65> "Aviation extension for countrycode"
[ifr]: <https://en.wikipedia.org/wiki/Instrument_flight_rules> "IFR Instrumenting Flight Rules"
[gmrepo]: <https://github.com/null2/globalmigration> "visualization of global migration flows' code repository"
[d3js]: <https://d3js.org/> "D3: Data Driven Documents"
[chorddia]: <https://bl.ocks.org/mbostock/4062006> "Chord diagram"
[circflows]: <https://github.com/espinielli/circflows> "circflows"